Is There a Competitive Advantage to Using Multivariate Statistical or Machine Learning Methods Over the Bross Formula in the hdPS Framework for Bias and Variance Estimation?

### 2024-09-24 Writeup/manuscript.tex
bmc correct

We observe that, aside from the Bross formula hdPS and Random Forest
(RF) models, where the number of proxies was manually set to 100, and
the kitchen sink (KS) model, which includes all proxies by design, the
number of proxy variables selected by other models ranges between 49 and
69. As expected, the hybrid method combining Bross and LASSO hdPS
selects exactly 49 proxy variables, all of which are selected by both
methods, resulting in a common feature rate of 1.00. For other models,
the common feature percentage is generally clustered around \(74\%\),
with XGBoost showing the highest common percentage at \(79\%\), while
the Genetic Algorithm (GA) displays the lowest common percentage at
\(69\%\).

\begin{table}[htbp]
\centering
\caption{Comparison of the count and percentage of proxy variables selected by each methods in common with that by the Bross formula hdPS}
\label{tab:proxy-common-pct-bross}
\begin{tabular}{lccc}
\toprule
\textbf{Method} & \textbf{Total Count} & \textbf{Common Count} & \textbf{Rate in Common} \\ 
\midrule
\textbf{Kitchen sink} & 142 & - & - \\
\textbf{Bross formula} & 100 & - & - \\
\textbf{Hybrid (Bross and LASSO)} & 49 & 49 & 1.00 \\
\textbf{LASSO} & 60 & 47 & 0.78 \\
\textbf{Elastic Net} & 69 & 54 & 0.78 \\
\textbf{Random Forest (RF)} & 100 & 71 & 0.71 \\  
\textbf{XGBoost} & 48 & 38 & 0.79 \\
\textbf{Forward selection (FS)} & 59 & 45 & 0.76 \\
\textbf{Backward elimination (BE)} & 59 & 45 & 0.76 \\
\textbf{Genetic algorithm (GA)} & 64 & 44 & 0.69 \\
\bottomrule
\end{tabular}
\end{table}

@@

Across the three scenarios---Frequent Exposure and Outcome, Rare
Exposure and Frequent Outcome, and Frequent Exposure and Rare
Outcome---XGBoost consistently exhibited the lowest Mean Squared Error
(MSE) and high coverage, making it one of the most reliable methods
overall. However, XGBoost did not always have the lowest bias; in fact,
methods such as the kitchen sink model, Bross-based hdPS, and Hybrid
hdPS often showed lower bias. The GA, in contrast, consistently
exhibited the highest bias and MSE, along with lower coverage and
greater variability, indicating it was less reliable for accurate effect
estimation. The kitchen sink model, while demonstrating low bias and
reasonable MSE, generally had higher variability and lower precision
compared to other methods like XGBoost. Methods such as Bross-based
hdPS, Hybrid hdPS, and Elastic Net performed moderately well across all
scenarios, balancing bias, coverage, and MSE, but they did not
outperform XGBoost in overall accuracy and precision.

@@

Scenarios with rare exposure tended to produce higher bias, particularly
for methods like GA and XGBoost, while frequent outcomes generally led
to lower bias. Overcoverage was common in the frequent exposure and
outcome scenario, while the other scenarios tended to have more balanced
or slightly under-coverage. Frequent exposure scenarios had higher
relative error, indicating more difficulty in precisely estimating
effects compared to rare exposure scenarios, which had lower relative
error. Rare outcome scenarios tended to have the lowest MSE, indicating
more precise effect estimates under these conditions, whereas rare
exposure scenarios had higher MSE, reflecting the challenges of
estimating effects when exposure is uncommon.

@@

\textcolor{red}{*need to write what hdps offers in the literature; what additional selection criteria was used so far*}

@@

\textcolor{red}{*summarize real data analysis finding in 1 para*}

@@

\textcolor{red}{*TMLE*}

@@

In conclusion, this analysis highlights the importance of carefully
selecting appropriate methods for hdPS analysis based on the specific
characteristics of the data, particularly the prevalence of exposure and
outcome. While XGBoost consistently demonstrated strong performance in
terms of MSE and coverage, it did not always have the lowest bias,
suggesting that it may be most suitable when precision is prioritized
over bias minimization. GA, despite its potential, showed significant
limitations with consistently high bias and MSE, making it less reliable
for accurate effect estimation. The kitchen sink model, Bross-based
hdPS, and Hybrid hdPS methods provided a balanced approach, often
delivering low bias and moderate MSE, but with variability in coverage
depending on the scenario. Scenario-specific trends revealed that rare
outcomes generally yielded lower MSE and better precision, while rare
exposures were associated with higher bias and MSE, emphasizing the
challenges of accurately estimating effects in such contexts.
Ultimately, the findings underscore the need to tailor method selection
to the epidemiological scenario at hand, ensuring that the chosen
approach aligns with the specific goals and challenges of the analysis.

@@

\begin{itemize}
\tightlist
\item
  hdPS: High-dimensional Propensity Score
\item
  NHANES: National Health and Nutrition Examination Survey
\item
  OR: Odds Ratio
\item
  RD: Risk Difference
\item
  SE: Standard Error
\item
  MSE: Mean Squared Error
\item
  \textcolor{blue}{KS: Kitchen Sink}
\item
  LASSO: Least Absolute Shrinkage and Selection Operator
\item
  \textcolor{blue}{ElasticNet / EN}: A regularized regression method
  that combines LASSO and Ridge regression
\item
  RF: Random Forest
\item
  XGBoost: Extreme Gradient Boosting
\item
  FS: Forward Selection
\item
  BE: Backward Elimination
\item
  GA: Genetic Algorithm
\item
  CV: Cross-Validation
\item
  RR: Relative Risk
\end{itemize}

@@

@@

@@

Over the past three years, MEK has received consulting fees from Biogen
Inc.~for consulting unrelated to this current work. MEK was previously
supported by the Michael Smith Foundation for Health Research Scholar
award.

@@

This work was supported by MEK's Natural Sciences and Engineering
Research Council of Canada (NSERC) Discovery Grants and Discovery
Accelerator Supplements.

>>>>

We observe that, aside from the Bross formula hdPS and Random Forest
models, where the number of proxies was manually set to 100, and the
kitchen sink (KS) model, which includes all proxies by design, the
number of proxy variables selected by other models ranges between 49 and
69. As expected, the hybrid method combining Bross and LASSO hdPS
selects exactly 49 proxy variables, all of which are selected by both
methods, resulting in a common feature rate of 1.00. For other models,
the common feature percentage is generally clustered around \(74\%\),
with XGBoost showing the highest common percentage at \(79\%\), while
the Genetic Algorithm (GA) displays the lowest common percentage at
\(69\%\).

\begin{table}[htbp]
\centering
\caption{Comparison of the count and percentage of proxy variables selected by each methods in common with that by the Bross formula-based high-dimensional propensity score to evaluate the association between obesity and diabetes from the National Health and Nutrition Examination Survey (NHANES) for the years 2013-2018.}
\label{tab:proxy-common-pct-bross}
\begin{tabular}{lccc}
\toprule
\textbf{Method} & \textbf{Total Count} & \textbf{Common Count} & \textbf{Rate in Common} \\ 
\midrule
\textbf{Kitchen sink} & 142 & - & - \\
\textbf{Bross formula} & 100 & - & - \\
\textbf{Hybrid (Bross and LASSO)} & 49 & 49 & 1.00 \\
\textbf{LASSO} & 60 & 47 & 0.78 \\
\textbf{Elastic Net} & 69 & 54 & 0.78 \\
\textbf{Random Forest} & 100 & 71 & 0.71 \\  
\textbf{XGBoost} & 48 & 38 & 0.79 \\
\textbf{Forward selection} & 59 & 45 & 0.76 \\
\textbf{Backward elimination} & 59 & 45 & 0.76 \\
\textbf{Genetic algorithm} & 64 & 44 & 0.69 \\
\bottomrule
\end{tabular}
\end{table}

@@

Across the three scenarios, XGBoost consistently achieved the lowest MSE
and high coverage, making it one of the most reliable methods in terms
of precision. However, it consistently exhibited some degree of bias,
particularly when compared to methods such as the kitchen sink model,
Bross-based hdPS, and Hybrid hdPS, which often showed lower bias in
scenarios with frequent outcomes. In contrast, GA displayed the highest
bias and MSE, along with lower coverage and greater variability, making
it the least reliable method for accurate effect estimation. Methods
such as Bross-based hdPS, Hybrid hdPS, and Elastic Net performed
moderately well across all scenarios, balancing bias, coverage, and MSE.
However, these methods did not outperform XGBoost in terms of overall
precision, especially with respect to MSE, though they often resulted in
lower bias. The kitchen sink model performed comparably to Bross-based
hdPS in terms of bias and coverage but lagged in SE estimation and MSE.

@@

In scenarios with rare exposure, higher bias was observed, particularly
for methods such as GA and XGBoost, whereas frequent outcomes generally
led to lower bias across most methods. Overcoverage was more common in
the scenario with frequent exposure and outcome, with several methods
producing confidence intervals that were too wide, suggesting an
overestimation of uncertainty. The other scenarios exhibited more
balanced or slightly under-coverage. Scenarios with frequent exposure
also displayed higher relative error in SE estimation, making it more
difficult to precisely estimate effects. In contrast, rare exposure
scenarios were associated with higher MSE, reflecting the challenge of
estimating effects when the exposure is uncommon. Rare outcome scenarios
exhibited the lowest MSE across most methods (except GA), indicating
that these scenarios provided a better balance between bias and
precision for most methods.

@@

Previous studies have shown that LASSO performs well within the hdPS
framework, particularly in terms of bias reduction and MSE
\citep{karim2018can, franklin2015regularized}, with similar results for
Elastic Net \citep{karim2018can}. Our findings largely support these
conclusions, as both methods demonstrated moderate bias and MSE across
the scenarios. However, their performance was inconsistent in some
cases, with higher bias observed in rare exposure scenarios, as noted in
previous studies \citep{karim2018can, franklin2015regularized}. MSE also
increased in rare exposure scenarios. The Hybrid hdPS method, combining
Bross-based hdPS and LASSO, showed promising results, especially in
terms of MSE, suggesting it may serve as a suitable alternative to
traditional hdPS methods. Random Forest also performed similarly in our
study, yielding relatively low bias, which is consistent with previous
findings \citep{karim2018can}. However, none of the earlier works
emphasized coverage or examined methods such as GA, XGBoost, forward
selection, or backward elimination, which were evaluated here.

@@

The real-world dataset, with frequent exposure (48.8\%) and a moderate
outcome rate (23.7\%), produced ORs of 1.59 and 1.56 for RF and XGBoost,
respectively, while other methods clustered around an OR of 1.52. A
general trend emerged, showing that methods selecting fewer proxy
variables yielded lower ORs, except for RF and the Bross formula hdPS.
In terms of risk difference (RD), estimates were relatively stable
across methods. Regarding variable overlap, most methods shared around
74\% of their proxy variables with the Bross formula hdPS, with XGBoost
showing the highest common rate (79\%) and GA the lowest (69\%).
Computing time analysis showed that, aside from RF and GA, most methods
had relatively fast computing times, with RF being significantly slower.

\subsection{Strengths}\label{strengths}

Previous studies on singly robust methods have mainly focused on hdPS
performance using MSE as the primary evaluation metric
\citep{franklin2015regularized, pang2016effect, wyss2018using, karim2018can, simon2023evaluating}.
Our study extends this body of work by incorporating a broader range of
performance metrics, allowing researchers to compare results in terms of
both bias and variance estimation. This comprehensive comparison of
statistical and machine learning methods across various scenarios has
not been conducted before. For instance, we found that while XGBoost
consistently demonstrated strong performance in terms of MSE and
coverage, it did not always have the lowest bias. Additionally, we
observed that traditional variable selection methods such as forward
selection and backward elimination performed similarly to more
sophisticated methods such as LASSO and Random Forest across scenarios,
both in terms of bias and coverage.

Our study also employed a complex plasmode simulation framework, closely
replicating real-world data conditions. This framework not only
accounted for model misspecification, where the true relationships
between covariates and outcomes were unknown, but also introduced noise
variables, addressing the challenge of dealing with irrelevant
covariates in high-dimensional settings. By applying transformations to
continuous variables and adding proxy variables that contributed
minimally to the outcome, we rigorously evaluated how well each method
handled both model uncertainty and non-informative variables, further
strengthening the real-world relevance of our findings.

@@

Double robust methods have demonstrated strong potential for achieving
optimal statistical performance in hdPS analyses
\citep{pang2016targeted, pang2016effect, benasseur2022comparison}. In
addition, single robust methods, such as ensemble learners such as super
learners, have shown promise in improving bias, MSE, area under the
curve (AUC), and covariate balance in other contexts
\citep{guertin2016head, wyss2018using, ju2019propensity}. Deep learning
methods, particularly supervised architectures, offer significant
potential for improving propensity score estimation in high-dimensional
settings by capturing complex, non-linear relationships and performing
well in scenarios with rare exposures, while maintaining comparable bias
and superior variance estimation compared to traditional methods
\citep{karim2024can}. Despite their theoretical advantages, the
complexity and computational demands of these methods, especially in
high-dimensional settings, have limited their adoption by practitioners.
On the other hand, the simpler singly robust machine learning methods
evaluated here have been applied in clinical research
\citep{hossain2023role, basham2021post}.

Future research should explore the application of double robust methods
and super learners in hdPS analyses, particularly for handling rare
outcomes and exposures, where the performance of traditional methods may
be suboptimal. Additionally, investigating the impact of different
hyperparameters on machine learning methods such as XGBoost could
optimize their performance in hdPS analysis. Finally, future simulation
studies should focus on evaluating coverage and other metrics in
epidemiological scenarios such as time-varying exposures, which would
provide valuable insights into the generalizability of our findings
\citep{neugebauer2015high}.

@@

In conclusion, this analysis highlights the importance of carefully
selecting appropriate methods for hdPS analysis based on the specific
characteristics of the data, particularly the prevalence of exposure and
outcome. These findings also emphasize the need to tailor method
selection to the specific epidemiological scenario, ensuring that the
chosen method aligns with the study's goals, whether minimizing bias or
maximizing precision.

XGBoost consistently demonstrated strong performance in terms of MSE and
coverage, making it an effective choice when precision is prioritized.
However, it did not achieve the lowest bias, particularly in rare
exposure scenarios, indicating that it may be less suited when
minimizing bias is the primary objective. In contrast, the Genetic
Algorithm (GA) exhibited significant limitations, with consistently high
bias and MSE, making it less reliable for effect estimation.

The kitchen sink, Bross-based hdPS, and Hybrid hdPS methods provided a
more balanced approach, offering low bias and moderate MSE, though
coverage varied depending on the scenario. The analysis also revealed
that rare outcomes were associated with lower MSE and better precision,
while rare exposures presented challenges, yielding higher bias and MSE.
Interestingly, traditional statistical approaches such as forward
selection and backward elimination performed comparably to more
sophisticated machine learning methods across many scenarios. This
suggests that simpler approaches can still be viable, particularly in
terms of bias and coverage, and might be preferred due to their
computational efficiency.

@@

\begin{itemize}
\tightlist
\item
  hdPS: High-dimensional Propensity Score
\item
  NHANES: National Health and Nutrition Examination Survey
\item
  OR: Odds Ratio
\item
  RD: Risk Difference
\item
  SE: Standard Error
\item
  MSE: Mean Squared Error
\item
  KS: Kitchen Sink
\item
  LASSO: Least Absolute Shrinkage and Selection Operator
\item
  EN: Elastic Net; a regularized regression method that combines LASSO
  and Ridge regression
\item
  RF: Random Forest
\item
  XGBoost: Extreme Gradient Boosting
\item
  FS: Forward Selection
\item
  BE: Backward Elimination
\item
  GA: Genetic Algorithm
\item
  CV: Cross-Validation
\item
  RR: Relative Risk
\end{itemize}

@@

The National Health and Nutrition Examination Survey (NHANES), conducted
by the U.S. Centers for Disease Control and Prevention (CDC), involves
collecting data through direct physical examinations, laboratory
testing, and interviews. The CDC already obtains consent from
participants when collecting this data. When researchers use NHANES data
for their studies, they are typically using de-identified, publicly
available data. This means that the information cannot be linked back to
individual participants, and therefore, additional consent from
participants is not required for researchers to use this data.

@@

NHANES data is publicly accessible and can be retrieved from the NHANES
website. The datasets generated and/or analyzed during the current study
are available in the NHANES repository:
\url{https://www.cdc.gov/nchs/nhanes/index.htm}. Additionally, all
simulation results can be interactively reviewed through a Shiny web
application at \url{https://ehsanx.shinyapps.io/hdPS-Alternatives/}.

\subsection*{Clinical trial number}\label{clinical-trial-number}
\addcontentsline{toc}{subsection}{Clinical trial number}

@@

MEK is currently supported by grants from Canadian Institutes of Health
Research and MS Canada. MEK has previously received consulting fees from
Biogen Inc.~for consulting unrelated to this current work. MEK was also
previously supported by the Michael Smith Foundation for Health Research
Scholar award.

@@

This work was supported by MEK's Natural Sciences and Engineering
Research Council of Canada (NSERC) Discovery Grant (PG\#: 20R01603) and
Discovery Launch Supplement (PG\#: 20R12709).
